\documentclass{article}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{trees, positioning}
\geometry{margin=1in}

\title{Binary Search Trees}
\author{}
\date{}

\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{cppstyle}{
  backgroundcolor=\color{backcolour},
  commentstyle=\color{codegray},
  keywordstyle=\color{blue},
  numberstyle=\tiny\color{codegray},
  stringstyle=\color{red},
  basicstyle=\ttfamily\footnotesize,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  keepspaces=true,
  numbers=none,
  numbersep=5pt,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2,
  language=C++
}

\begin{document}

\maketitle

A \textbf{Binary Search Tree (BST)} is a binary tree where each node contains a key such that:
\[
\texttt{left subtree keys} < \texttt{node key} < \texttt{right subtree keys}
\]
This ordering enables efficient searching, insertion, and deletion operations.

Similar to hashtables, they are used to store key-value pairs. However unlike hashtables, BSTs maintain keys in sorted order.


\section{Example BST Structure}

\begin{center}
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=3.5cm},
  level 2/.style={sibling distance=2cm}
]
\node {8}
  child {node {3}
    child {node {1}}
    child {node {6}
      child {node {4}}
      child {node {7}}
    }
  }
  child {node {10}
    child[missing]
    child {node {14}
      child {node {13}}
      child {node {null}}
    }
  };
\end{tikzpicture}
\end{center}

This BST contains keys ordered such that for every node:
\[
\texttt{left subtree keys} < \texttt{node key} < \texttt{right subtree keys}
\]

\section{Use Cases}
\begin{itemize}
  \item Symbol tables in compilers
  \item Maintaining sorted datasets
  \item Searching and dynamic set operations
\end{itemize}

\section{Time Complexity}

\begin{tabular}{|c|c|c|}
\hline
\textbf{Operation} & \textbf{Average Case} & \textbf{Worst Case (Unbalanced)} \\
\hline
Search & $\mathcal{O}(\log n)$ & $\mathcal{O}(n)$ \\
Insert & $\mathcal{O}(\log n)$ & $\mathcal{O}(n)$ \\
Delete & $\mathcal{O}(\log n)$ & $\mathcal{O}(n)$ \\
\hline
\end{tabular}

\subsection{Terminology}
\begin{itemize}
  \item \textbf{BST:} A binary tree with ordered keys
  \item \textbf{Leaf:} A node with no children
  \item \textbf{Internal node:} Has at least one child
  \item \textbf{Height:} Number of edges on the longest downward path
  \item \textbf{Balanced Tree:} Height is approximately $\log n$
\end{itemize}



\section{Why BST Operations Are $\mathcal{O}(\log n)$}

In a \textbf{balanced} BST, each level of the tree halves the search space. With $n$ nodes, the maximum height is approximately $\log_2 n$.

\begin{itemize}
  \item At each comparison, we eliminate half the remaining nodes.
  \item This is analogous to binary search on a sorted array.
\end{itemize}

Hence, insertion, search, and deletion all take $\mathcal{O}(\log n)$ time on average for balanced trees.



\subsection{C++ Implementation of a BST}

\begin{lstlisting}[style=cppstyle]
#include 
using namespace std;

struct BSTNode {
  int key;
  BSTNode* left;
  BSTNode* right;

  BSTNode(int val) : key(val), left(nullptr), right(nullptr) {}
};

class BST {
public:
  BST() : root(nullptr) {}

  void insert(int key) {
    root = insert(root, key);
  }

  bool search(int key) const {
    return search(root, key);
  }

  int min() const {
    BSTNode* node = min(root);
    if (node) return node->key;
      throw runtime_error("Tree is empty");
  }

  int max() const {
    BSTNode* node = max(root);
    if (node) return node->key;
      throw runtime_error("Tree is empty");
  }

private:
  BSTNode* root;

  BSTNode* insert(BSTNode* node, int key) {
    if (!node) return new BSTNode(key);
    if (key < node->key)
      node->left = insert(node->left, key);
    else if (key > node->key)
      node->right = insert(node->right, key);
    return node;
  }

  bool search(BSTNode* node, int key) const {
    if (!node) return false;
    if (key == node->key) return true;
    if (key < node->key)
      return search(node->left, key);
    else
      return search(node->right, key);
  }

  BSTNode* min(BSTNode* node) const {
    if (!node) return nullptr;
    while (node->left)
        node = node->left;
    return node;
  }

  BSTNode* max(BSTNode* node) const {
    if (!node) return nullptr;
    while (node->right)
        node = node->right;
    return node;
  }

};
\end{lstlisting}

\textbf{Inorder Traversal (Left → Root → Right):}
\begin{lstlisting}[style=cppstyle]
void inorder(BSTNode* node) const {
  if (!node) return;
  inorder(node->left);
  cout << node->key << " ";
  inorder(node->right);
}
\end{lstlisting}

\textbf{Preorder Traversal (Root → Left → Right):}
\begin{lstlisting}[style=cppstyle]
void preorder(BSTNode* node) {
  if (!node) return;
  cout << node->key << " ";
  preorder(node->left);
  preorder(node->right);
}
\end{lstlisting}

\textbf{Postorder Traversal (Left → Right → Root):}
\begin{lstlisting}[style=cppstyle]
void postorder(BSTNode* node) {
  if (!node) return;
  postorder(node->left);
  postorder(node->right);
  cout << node->key << " ";
}
\end{lstlisting}

\section{Deletion in BSTs}

To delete a node, we consider three cases:

\begin{itemize}
  \item \textbf{Case 1: Leaf Node (0 children)}  
    Simply remove the node.
  \item \textbf{Case 2: One Child}  
    Replace the node with its only child.
  \item \textbf{Case 3: Two Children}  
    Replace the node’s value with either:
    \begin{itemize}
      \item The \textbf{inorder successor} (smallest value in right subtree)
      \item The \textbf{inorder predecessor} (largest value in left subtree)
    \end{itemize}
    Then delete the successor or predecessor node.
\end{itemize}

\subsubsection*{Finding the Inorder Successor}

To find the inorder successor of a node in a BST, we consider two cases:

\begin{itemize}
  \item If the node has a right subtree: successor is the leftmost node in the right subtree
  \item If the node has no right subtree: successor is the lowest ancestor for which the node lies in its left subtree
\end{itemize}

\subsection{Visualizing Inorder Successor Cases}

\begin{center}
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=3.5cm},
  level 2/.style={sibling distance=2cm},
  level 3/.style={sibling distance=1.5cm},
  successor/.style={circle, draw, minimum size=0.8cm, thick, color=blue},
  target/.style={circle, draw, minimum size=0.8cm, thick, color=red}
]

% Tree structure
\node[successor] (n8) {8}
  child {node {3}
    child {node {1}}
    child {node[target] (n6) {6}
      child {node {4}}
      child {node[target] (n7) {7}}
    }
  }
  child {node {10}
    child[missing]
    child {node {14}}
  };

% Legend
\node[rectangle, draw, minimum width=3.2cm, minimum height=1cm, right=3.5cm of n8] (legend) {
  \begin{tabular}{ll}
  \textcolor{red}{\textbf{Red}} & Target Node \\
  \textcolor{blue}{\textbf{Blue}} & Inorder Successor
  \end{tabular}
};

\end{tikzpicture}
\end{center}

\textbf{Case 1:}  
Node \textbf{6} has a right child (7). Its inorder successor is the leftmost node in its right subtree: \textbf{7}.



\textbf{Case 2:}  
Node \textbf{7} has no right subtree. To find its inorder successor:

\begin{itemize}
  \item We move up the tree from node 7 to its parent (node 6), but since 7 is a \textbf{right child}, we continue upward.
  \item We then move from 6 to its parent (node 3), where 6 is a \textbf{right child} again, so we continue.
  \item We then move from 3 to node 8, where 3 is a \textbf{left child}.
  \item Hence, node \textbf{8} is the lowest ancestor where node 7 (through its lineage) lies in the left subtree.
\end{itemize}

Therefore, the inorder successor of node \textbf{7} is \textbf{8}.


\subsection{Degenerate Cases and Balancing}

A poorly structured BST can degrade to a linked list, resulting in $\mathcal{O}(n)$ time operations. To prevent this:

\begin{itemize}
  \item Use self-balancing trees like AVL or Red-Black Trees
  \item Randomize insertions or rebalance periodically
\end{itemize}

Balanced BSTs ensure height $\mathcal{O}(\log n)$, which guarantees logarithmic performance for search-related operations.

\section{Tree Rotations}

Tree rotations are fundamental operations used in self-balancing binary search trees (e.g., AVL and Red-Black Trees) to restore height balance after insertions or deletions. A rotation is a local restructuring operation that preserves the BST ordering.

\subsection{Right Rotation (\texttt{rotateRight})}

Right rotation is applied at a node $y$ with a left child $x$ to reduce the height of the left subtree.

\textbf{Before:}
\[
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=2.5cm}
]
\node (y) {y}
  child {node (x) {x}
    child {node {A}}
    child {node {B}}
  }
  child {node {C}};
\end{tikzpicture}
\]

\textbf{After \texttt{rotateRight(y)}:}
\[
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=2.5cm}
]
\node (x) {x}
  child {node {A}}
  child {node (y) {y}
    child {node {B}}
    child {node {C}}
  };
\end{tikzpicture}
\]

\subsection{Left Rotation (\texttt{rotateLeft})}

Left rotation is applied at a node $x$ with a right child $y$ to reduce the height of the right subtree.

\textbf{Before:}
\[
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=2.5cm}
]
\node (x) {x}
  child {node {A}}
  child {node (y) {y}
    child {node {B}}
    child {node {C}}
  };
\end{tikzpicture}
\]

\textbf{After \texttt{rotateLeft(x)}:}
\[
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=2.5cm}
]
\node (y) {y}
  child {node (x) {x}
    child {node {A}}
    child {node {B}}
  }
  child {node {C}};
\end{tikzpicture}
\]

\subsection{Preserving the BST Property During Rotations}

Tree rotations maintain the binary search tree property:
\[
\texttt{left subtree} < \texttt{node} < \texttt{right subtree}
\]

Consider the right rotation at node $y$ (where $x$ is its left child):

\textbf{Before rotation:}
\begin{itemize}
  \item Subtree A: All keys $<$ x.key
  \item Subtree B: x.key $<$ keys $<$ y.key
  \item Subtree C: All keys $>$ y.key
\end{itemize}

\textbf{Right Rotation:}

\[
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=2.5cm}
]
\node (y) {y}
  child {node (x) {x}
    child {node {A}}
    child {node {B}}
  }
  child {node {C}};
\end{tikzpicture}
\quad
\Rightarrow
\quad
\begin{tikzpicture}[
  every node/.style={circle, draw, minimum size=0.8cm},
  level distance=1.4cm,
  level 1/.style={sibling distance=2.5cm}
]
\node (x) {x}
  child {node {A}}
  child {node (y) {y}
    child {node {B}}
    child {node {C}}
  };
\end{tikzpicture}
\]

\textbf{After rotation:}
\begin{itemize}
  \item Subtree A remains in the left of x
  \item Subtree B becomes left of y, and since B's keys were $>$ x.key and $<$ y.key, they remain valid
  \item Subtree C remains to the right of y
\end{itemize}

All relative key orderings are preserved, satisfying:
\[
\texttt{A} < x < \texttt{B} < y < \texttt{C}
\]


Both left and right rotations preserve in-order traversal and relative ordering of keys. Hence, the BST property remains intact after any single rotation.


\section{std::map and std::set in C++ STL}

The C++ Standard Template Library (STL) provides \verb|std::map| and \verb|std::set| as associative containers that store elements in sorted order.

\verb|std::map| stores key-value pairs where keys are unique and each key maps to a corresponding value.  
\verb|std::set| stores unique keys without associated values.

Both containers are typically implemented using a balanced binary search tree, most commonly a \emph{Red-Black Tree}. This ensures that insertions, deletions, and lookups have logarithmic time complexity: $O(\log n)$.

Key properties:
\begin{itemize}
  \item Elements are kept sorted according to the comparison function (default: \verb|operator<|).
  \item Iteration over \verb|map| or \verb|set| yields elements in sorted order.
  \item Efficient operations such as range queries, lower and upper bound searches, and equal range queries are supported.
\end{itemize}

\paragraph{Relation to Binary Search Trees.}  
\verb|std::map| and \verb|std::set| are abstracted interfaces over balanced binary search trees. These containers provide the ordered behavior and logarithmic performance characteristics of binary search trees while hiding the tree operations from the user.

\paragraph{Contrast with Hash-Based Containers.}  
Unlike \verb|std::unordered_map| and \verb|std::unordered_set|, which use hash tables:
\begin{itemize}
  \item \verb|map| and \verb|set| maintain elements in order.
  \item Lookup, insertion, and deletion are $O(\log n)$ rather than average $O(1)$ (hash containers) but no ordering.
  \item \verb|map| and \verb|set| support ordered range operations; hash containers do not.
\end{itemize}


\section{Practice problem}
Leetcode Problem 108: \textit{Convert Sorted Array to Binary Search Tree}

\url{https://leetcode.com/problems/convert-sorted-array-to-binary-search-tree/description/}

\end{document}